---
title: "Data Exploration of 2015 Flight Data and TSA Claims"
output:
  html_document: default
  html_notebook: default
  df_print: paged
  pdf_document: default
  
---

```{r include=FALSE, cache=FALSE}
library(scales)
library(grid)
library(RColorBrewer)

library(tidyverse)
library(lubridate) # Part of the Tidyverse to handle dates and times
library(dplyr)

library(readxl)
library(stringr)
library("ggpubr")
library("readr")
library("plotly")
setwd(dir = "C:/Users/rknaa/Documents/datascience/dataexploration-final-project/")
```
# Flight Data
```{r warning=FALSE}
#small for testing 

#totflights <- read_csv(file='totflights-small.csv')
totflights <- read_csv(file='totflights.csv')


#load("totflights.Rda")
#structure(totflights)

# mutate columns to factors, certain functions work with columns as factors
totflights <-totflights %>% mutate(CARRIER = factor(CARRIER))
totflights <-totflights %>% mutate(DEST_AIRPORT_ID = factor(DEST_AIRPORT_ID))
totflights <-totflights %>% mutate(ORIGIN_AIRPORT_ID = factor(ORIGIN_AIRPORT_ID))
totflights <-totflights %>% mutate(MONTH = factor(MONTH))

#save(totflights,file="totflights.Rda")

#save(totflights,file="totflights.Rda")
#rm(totflights)

#totflights
```
Summary of total flight data for 2015.
```{r}
summary(totflights)
```

```{r}
# The main data from flights I am concerned about is Departure Delay 
min(totflights$DEP_DELAY)

# Let's find the mean departure delay
mean(totflights$DEP_DELAY)

#let's find the standard deviation
sd(totflights$DEP_DELAY)

#most departing flights by airport
flight_count_by_airport <- totflights%>% group_by(ORIGIN_AIRPORT_ID) %>% summarise(count =n()) %>% arrange(desc(count))
cat (toString(flight_count_by_airport$ORIGIN_AIRPORT_ID[1]), 'had the most flights with',flight_count_by_airport$count[1], 'flights.')

#flight with the longest departure day
longest_departure_delay <- totflights %>% arrange(desc(DEP_DELAY))
#longest_departure_delay

cat (toString(longest_departure_delay$ORIGIN_AIRPORT_ID[1]) , 'to',toString(longest_departure_delay$DEST_AIRPORT_ID[1]),'on',format(longest_departure_delay$FL_DATE[1],"%m-%d-%Y"),'had the longest delay of',longest_departure_delay$DEP_DELAY[1], 'minutes.')

```


Some basic visualizations of my data of interest.
```{r}
hist(totflights$DEP_DELAY)
hist(totflights$ARR_DELAY)
```


Check for normality of Depature Delay or Arrival Delay

```{r}
#check for normality
ggqqplot(totflights$DEP_DELAY, ylab = "Departure Delay",title='QQ Plot Departure Delay')
ggqqplot(totflights$ARR_DELAY, ylab = "Arrival Delay",title='QQ Plot Arrival Delay')

```

```{r warning=FALSE}

#A lot of airports with a small number of flights.  Sample size is going to be too small 

# Let's try adding a filter so that only airports with more than 5,000 flights in a year are represented

delaysbyairport <- totflights %>% group_by(ORIGIN_AIRPORT_ID) %>% summarise(count =n(),delay = mean(DEP_DELAY)) %>% filter(count > 5000) %>% arrange(desc(delay))


#much better,  large airports and acceptable seeming delays

#plot a slice of the dataframe to keep the chart sane

delaysbyairport %>% slice(1:10)

delayplot <-ggplot(data=delaysbyairport %>% slice(1:10), aes(x = ORIGIN_AIRPORT_ID, y = delay,fill=ORIGIN_AIRPORT_ID)) +
  geom_bar(stat="identity") +
  labs(title = "Highest Average Departure Delay 2015",
       y = "Minutes",
       x = "Airport") + theme_bw(base_size = 15)  + theme(legend.position="none", 
                                                          panel.grid.major = element_blank(),
                                                          panel.grid.minor = element_blank(),
                                                          panel.border = element_blank(),
                                                          panel.background = element_blank())

delayplot


```


Now let's look at delays on a monthly basis
```{r}
#Let's look at delays per month

head(totflights)

months <-levels(totflights$MONTH)
avg_delays_by_month=c()

for(filtermonth in months){
  #print(filtermonth)
  temp_monthly_flights <- totflights %>% filter(MONTH == filtermonth) %>%  summarise(
     TOT_ARR_DELAY = sum(ARR_DELAY), MEAN_ARR_DELAY = mean(ARR_DELAY),
     TOT_DEP_DELAY = sum(DEP_DELAY), MEAN_DEP_DELAY = mean(DEP_DELAY), NUM_FLIGHTS  =(count=n())
  )
  temp_monthly_flights$month <- filtermonth
  avg_delays_by_month <- bind_rows(avg_delays_by_month,temp_monthly_flights)
  rm(temp_monthly_flights)
}

avg_delays_by_month %>% arrange(MEAN_DEP_DELAY)

#Print the month with the most claims and the ammount

cat (month.name[as.numeric(avg_delays_by_month$month[1])], 'had the shortest mean departure delay of',avg_delays_by_month$MEAN_DEP_DELAY[1], 'minutes.')

cat (month.name[as.numeric(avg_delays_by_month$month[length(avg_delays_by_month$month)])], 'had the longest mean departure delay of',avg_delays_by_month$MEAN_DEP_DELAY[length(avg_delays_by_month$MEAN_DEP_DELAY)], 'minutes.')



monthly_data <- data.frame(x <- month.abb[as.numeric(avg_delays_by_month$month)], y1 <-avg_delays_by_month$MEAN_DEP_DELAY, y2 <-avg_delays_by_month$MEAN_ARR_DELAY, y3 <-avg_delays_by_month$NUM_FLIGHTS)
monthly_flights_plot <- plot_ly(monthly_data, x = ~x, y = ~y3, type = 'bar', name = 'Flights', marker = list(color = 'rgb(49,130,189)')) %>%
  layout(title = "Number of Flights by Month",xaxis = list(title = "", tickangle = -45,categoryorder = "array",categoryarray = monthly_data$x),
         yaxis = list(title = ""),
         margin = list(b = 100),
         barmode = 'group')

monthly_flights_plot

monthly_delays_plot <- plot_ly(monthly_data, x = ~x, y = ~y1, type = 'bar', name = 'Departure Delay', marker = list(color = 'rgb(49,130,189)')) %>%
  add_trace(y = ~y2, name = 'Arrival Delay', marker = list(color = 'rgb(204,204,204)')) %>%
  layout(title = "Mean Delays in Minutes",xaxis = list(title = "", tickangle = -45,categoryorder = "array",categoryarray = monthly_data$x),
         yaxis = list(title = ""),
         margin = list(b = 100),
         barmode = 'group')

monthly_delays_plot


```





#TSA Claims

Next, I will explore the TSA Claims Data.
Start by reading in the CSV data creating a column for just the month, and filtering for claims that occurred in 2015.
```{r cache=FALSE}
#Start by read the data from the CSV, columns are rename for convenience
#TSA claims data
claims<- read_excel('claims-data-2015-as-of-feb-9-2016.xlsx')

colnames(claims)
claims <- rename (claims, claimnumber = "Claim Number", datereceived ="Date Received",
                  incidentd = "Incident D", airportcode = "Airport Code", airportname="Airport Name",
                  airlinename = "Airline Name", claimtype="Claim Type", claimsite ="Claim Site",
                  itemcategory = "Item Category", closeamount ="Close Amount", disposition = "Disposition")
claims <- claims %>% mutate(yearmonth = format(incidentd, "%Y-%m"), year = format(incidentd, "%Y"),month = format(incidentd, "%m"))
claims$month <- as.numeric(claims$month)

claims <- claims %>% filter(year == '2015')
```
Summary of claims.
```{r}
summary(claims)
```
Basic visualizations
```{r}
hist(claims$month)
#View(claims)
```
Why is January so high?
Let's get a count by month.
```{r}

maxclaims <- claims %>% group_by(month) %>% summarize(count = n()) %>% arrange(desc(count))

g <- ggplot(claims, aes(month)) 
g + geom_bar(fill="#c0392b")+scale_x_continuous(breaks=claims$month,labels=paste(month.abb[claims$month])) +
    labs(x="Month",title="Claims per month") 
```

```{r}
#Print the month with the most claims and the amount

cat (month.name[maxclaims$month[1]], 'had the most claims with',maxclaims$count[1], 'claims.')
```

```{r}
# Let's make a dataframe of airport codes and claims count, and then make a bar chart.
claimsbyairport <- claims %>% group_by(airportcode) %>% summarise(count =n()) %>% arrange(desc(count))
#claimsbyairport

# The main data from claimsbyairport I am concerned about is number of claims per airport 
min(claimsbyairport$count)

# Let's find the mean claims count 
mean(claimsbyairport$count)

#let's find the standard deviation
sd(claimsbyairport$count)

# Top ten claims count by airport
claimsbyairport %>% slice(1:10)

#plot a slice of the dataframe to keep the chart sane
claimsplot <-ggplot(data=claimsbyairport %>% slice(1:10), aes(x = airportcode, y = count, fill= airportcode)) +
  geom_bar(stat="identity") +
  labs(title = "Top 10 Airports with most TSA Claims",
       y = "Claims",
       x = "Airport") + theme_bw(base_size = 15)  + theme(legend.position="none", panel.grid.major = element_blank(),
                                                          panel.grid.minor = element_blank(),
                                                          panel.border = element_blank(),
                                                          panel.background = element_blank())
claimsplot
```

#Combining the data sets
```{r warning=FALSE}
#There are a lot data points in the flight data set
#I think we would do better with just totals for each airport for the year

#Let's look at how many flights with delays there are for each Origin Airport, total delay time and mean delay time
total_dep_delays_by_airport <- totflights %>%
  group_by(ORIGIN_AIRPORT_ID) %>% filter(DEP_DELAY > 0) %>% summarise(TOT_DEP_DELAY = sum(DEP_DELAY), MEAN_DEP_DELAY = mean(DEP_DELAY), NUM_FLIGHTS  =(count=n()))
  
#Sanity check the results
glimpse(total_dep_delays_by_airport)
#total_dep_delays_by_airport


summary(total_dep_delays_by_airport$TOT_DEP_DELAY)
summary(total_dep_delays_by_airport$MEAN_DEP_DELAY)

#now let's build a list of claims for each airport

total_claims_by_airport = claims %>% group_by(airportcode) %>% summarise (total_claims = n())

total_claims_by_airport


# Join the flight data and claims data on airport code
combined_claims_delay <- full_join(total_dep_delays_by_airport,total_claims_by_airport,by = c('ORIGIN_AIRPORT_ID' = 'airportcode'))



```

# Analysis of Combined Data Sets
```{r}
#Let's try to answer our initial research question
#Is the number of TSa Claims at an airport corrolated to the the average delay time at that airport.

combined_claims_delay <-combined_claims_delay %>% filter(total_claims > 1)  
head(combined_claims_delay )

res <- cor.test(combined_claims_delay$MEAN_DEP_DELAY, combined_claims_delay$total_claims, method = "pearson")
print (res)
```
```{r}
cat ('With a correlation coefficient of',res$estimate, 'and a p-value of',res$p.value)
cat ('There is very weak negative correlation between TSA Claims and Mean Departure Delays' )
```

```{r warning=FALSE}
ggscatter(combined_claims_delay, x = "total_claims", y = "MEAN_DEP_DELAY",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "TSA Claims", ylab = "Mean Departure Delay",
          color='blue'
)


```

#Addtional possible correlations

Number of Flights as a predictor of TSA Claims
```{r warning=FALSE}
# Since we have all this data processed.  Let's look at addtional possible correlations

# Join the flight data and claims data on airport code
combined_claims_flight_count <- full_join(flight_count_by_airport,total_claims_by_airport,by = c('ORIGIN_AIRPORT_ID' = 'airportcode'))
combined_claims_flight_count
res <- cor.test(combined_claims_flight_count$total_claims, combined_claims_flight_count$count,method = "pearson")


cat ('With a correlation coefficient of',res$estimate, 'and a p-value of',res$p.value)
cat ('There is a postive correlation between number of flights and the number of TSA Claims' )

print (res)

combined_claims_flight_count_plot <- ggscatter(combined_claims_flight_count, x = "count", y = "total_claims",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Departure Flight Count", ylab = "TSA Claims",
          color='blue'
)
combined_claims_flight_count_plot


```


Mean departure delay as a predictor of arrival delays at the same airport.
```{r}
# Mean Arrival Delays and Mean Departure Delays 
airport_dep_delays <- totflights %>%
  group_by(ORIGIN_AIRPORT_ID) %>% summarise(TOT_DEP_DELAY = sum(DEP_DELAY), MEAN_DEP_DELAY = mean(DEP_DELAY)) 


airport_arr_delays <- totflights %>%
  group_by(DEST_AIRPORT_ID) %>% summarise(TOT_ARR_DELAY = sum(ARR_DELAY), MEAN_ARR_DELAY = mean(ARR_DELAY))

airport_dep_arr_delays <- full_join(airport_dep_delays,airport_arr_delays,by = c('ORIGIN_AIRPORT_ID' = 'DEST_AIRPORT_ID'))

airport_dep_arr_delays

res <- cor.test(airport_dep_arr_delays$MEAN_ARR_DELAY, airport_dep_arr_delays$MEAN_DEP_DELAY, method = "pearson")
print (res)

cat ('With a correlation coefficient of',res$estimate, 'and a p-value of',res$p.value)
cat ('There is postive correlation between Mean Arrival Delays and Mean Departure Delays' )

ggscatter(airport_dep_arr_delays, x = "MEAN_DEP_DELAY", y = "MEAN_ARR_DELAY",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Mean Arrival Delay", ylab = "Mean Departure Delay",
          color='blue'
)


```


